---
title: "Charlotte Lynx Blue Line's Impact on Air Pollution"
author: "Samirah Huda and Landon Thomas"
format: gfm
editor: visual
---

## Research Question

What effect does the Lynx Blue Line light rail have on urban air pollution in Charlotte, South Carolina?

What effect does the Lynx Blue Line light rail have on urban air pollution in Charlotte, South Carolina?

## Background

## Data

**PM2.5 Data from SEDAC**

We obtained daily PM2.5 concentrations for the contiguous United States at 1 km spatial resolution for the years 2000 to 2016 in GeoTIFF format from NASA's Socioeconomic Data and Applications Center (SEDAC). We then found the coordinates of the city center for the target city and each control city, and we created a buffer with a 10 kilometer radius around each city center. Finally, we cropped and adjusted the projections of the GeoTIFF files and extracted the PM2.5 data, creating a CSV file of the data set containing the monthly average PM2.5 concentrations for each city.

Landon explains work here

## Estimation Methodology & Results

**Machine Learning to Compare Predicted Baseline with Actual Data**

We used data before Charlotte's light rail opening to predict future PM2.5 levels had the light rail not opened. We then compared the predicted PM2.5 data with actual data after the light rail opening data. This comparison allowed us to gain more insight about whether or not the PM2.5 concentrations would have been higher if the light rail had not opened.

Machine learning methods we used include the Random Forest model and the XGBoost model.

Create training data

```{r}
#| warning: false
#| message: false
#install.packages("rmarkdown")
#install.packages("terra")
#install.packages("tidyverse")
#install.packages("dplyr")
library("tidyverse")
library("dplyr")
library("leaps")
library("rpart")
library("rpart.plot")
library("vip")
library("ranger")
library("xgboost")


updated <- read.csv("merged_pm25.csv")
```

# adjust data for machine learning models

```{r}
updated2<-updated %>%
  mutate(city=as.factor(city))%>%
  dplyr::select(-temp, -date, -meanpm25)
```

# Machine Learning Models:

# Upload and clean data

```{r}
df2<-updated2 %>%
  filter(lr_month<=36 & lr_month>= -36)

df3<-df2 %>%
  filter(lr_op < 1) %>%
  dplyr::select(-lr_op, -lr_month)

sapply(lapply(df2, unique), length)
```

# Split data in to test and train group

We used data before Charlotte's light rail opening to predict future PM2.5 levels had the light rail not opened. We then compared the predicted PM2.5 data with actual data after the light rail opening data. This comparison allowed us to gain more insight about whether or not the PM2.5 concentrations would have been higher if the light rail had not opened.

Machine learning methods we used include the Random Forest model and the XGBoost model.

Create training data

```{r}
library("tidyverse")
library("dplyr")
library("leaps")
library("rpart")
library("rpart.plot")
library("vip")
library("ranger")
library("xgboost")

updated <- read.csv("merged_pm25.csv")

# adjust data for machine learning models
updated2<-updated %>%
  mutate(city=as.factor(city))%>%
  dplyr::select(-temp, -date, -meanpm25)

# Machine Learning Models:

# Upload and clean data
df2<-updated2 %>%
  filter(lr_month<=36 & lr_month>= -36)

df3<-df2 %>%
  filter(lr_op < 1) %>%
  dplyr::select(-lr_op, -lr_month)

sapply(lapply(df2, unique), length)

set.seed(112)

shuffled<-df3 %>% sample_frac(size=1, replace=FALSE)

train<-shuffled %>%
  dplyr::slice(1:100)

test<-shuffled %>%
  dplyr::slice(101:145)
```

Estimate RMSE from the Random Forest Model

```{r}
model4 <- ranger(new_monthly_avg ~ ., data=train, importance='impurity')
v1 <- vip(model4)
v1
```

In this chart, the importance represents how heavily each variable influences the model. The higher the importance, the higher the impact that variable has. Here, the variable that most heavily influenced the results is the city. The second most influential variable is population, the third most is vapor pressure squared, and so on.

```{r}
#rf model evaluation
predict_test<-predict(model4, test)

rmse_test_rf<-sqrt(mean(test$new_monthly_avg-predict_test$predictions)^2)

rmse_test_rf

#create data for the graph
openlr <- updated %>%
  filter(lr_op>0) %>%
  mutate(city=as.factor(city))%>%
  dplyr::select(-temp, -date, -meanpm25) %>%
  dplyr::select(-lr_op, -lr_month)

# rf model with all data
predict_test<-predict(model4, openlr)

rf_pred<-predict_test$predictions
rf_df<-cbind(openlr, rf_pred)
```

Estimate RMSE with the XGBoost model

```{r}
# Split data in to test and train group
set.seed(112)

shuffled<-df3 %>% sample_frac(size=1, replace=FALSE)
all_data <- df2 %>%
  dplyr::select(-lr_op)
all_data2 <- data.matrix(all_data)

train<-shuffled %>%
  dplyr::slice(1:100)

test<-shuffled %>%
  dplyr::slice(101:145)

#define predictor and response variables in training set
train_x = data.matrix(train[, -4])
train_y = train[,4]

#define predictor and response variables in testing set
test_x = data.matrix(test[, -4])
test_y = test[, 4]

#define final training and testing sets
xgb_train = xgb.DMatrix(data = train_x, label = train_y)
xgb_test = xgb.DMatrix(data = test_x, label = test_y)

watchlist = list(train=xgb_train, test=xgb_test)

#fit XGBoost model
xgb_model = xgboost(data = xgb_train, max.depth = 3, nrounds = 14, verbose = 0)
predict_xgb<-predict(xgb_model, xgb_test)
rmse_test_xgb<-sqrt(mean(test_y-predict_xgb)^2)

rmse_test_xgb

#create data for the graph
openlr <- updated %>%
  filter(lr_op>0) %>%
  mutate(city=as.factor(city))%>%
  dplyr::select(-temp, -date, -meanpm25) %>%
  dplyr::select(-lr_op, -lr_month)

openlr_matrix <- data.matrix(openlr)

all_x = data.matrix(openlr_matrix[, -4])
all_y = openlr_matrix[, 4]

xgb_all = xgb.DMatrix(data = all_x, label = all_y)

# rf model with all data
predict_all_xgb<-predict(xgb_model, xgb_all)

xgb_df<-cbind(rf_df, predict_all_xgb)
```

Plot predicted PM2.5 with actual data

```{r}
openlr <- updated %>%
  filter(lr_op>0)

lr_removed <- df2 %>%
  dplyr::select(-lr_op)

# rf model with all data
model5 <- ranger(new_monthly_avg ~ ., data=lr_removed, importance='impurity')
predict_test<-predict(model4, lr_removed)

# xgb model with all data
matrix <- data.matrix(lr_removed)

my <- df2 %>%
  dplyr::select(lr_month, month, year, city)

pred_df<-merge(xgb_df, my, by=c("month","year", "city"), all.x=TRUE)

p <- ggplot() + 
  geom_line(data = df2, aes(x = lr_month, y = new_monthly_avg, color = "Actual Data"), linewidth=1) + facet_wrap(~ city) +
  geom_line(data=pred_df, aes(x = lr_month, y = rf_pred, color = "RF Counterfactual"), linewidth=1) + facet_wrap(~ city) +
  geom_line(data=pred_df, aes(x = lr_month, y = predict_all_xgb, color = "XGB Counterfactual"), linewidth=1) + facet_wrap(~ city) +
  xlab("Months since light rail opening") + ylab("Mean PM2.5 (μg/m3)") + geom_vline(xintercept=0, linetype="dashed")+theme_bw()

p
```

This plot compares the actual concentrations of PM2.5, and the predicted levels of PM2.5 predicted by both the Random Forest model and the XGBoost model, before and after light rails opened in Charlotte, Houston, Minneapolis, and Phoenix.

**Comparing PM2.5 Changes with Counterfactual Cities**

Landon work on this part

## Conclusion

Based on the machine learning models, the higher the RMSE (error squared), the less accurate the model is. In this study, the Random Forest Model produced a RMSE value of 0.2596, while the XGBoost model produced a RMSE value of 0.4430. Therefore, in this study, the Random Forest Model proved to be more accurate for predicting PM2.5 concentrations in these four cities had light rails not opened.

## References

Cite papers in APA format

## 
